### 2023-04-03 paper-thesis/chapters/000-abstract.tex
First part of feedback from Jesper

\noindent Static type systems can greatly enhance the quality of programs, but implementing a type checker is challenging and error-prone. The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system. However, so far Statix has not been used to implement a type system with dependent types, an expressive class of type systems which require evaluation of terms during type checking.

In this thesis, we present an specification of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments. While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.

>>>>

\noindent Static type systems can greatly enhance the quality of programs, but implementing a type checker for them is challenging and error-prone. The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system. However, so far Statix has not been used to implement a type system with dependent types, an expressive class of type systems which require evaluation of terms during type checking.

In this thesis, we present a specification of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments. While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.

### 2023-04-03 paper-thesis/chapters/01-introduction.tex
First part of feedback from Jesper

While we keep building more and more complex programming languages, their type checkers are still often written in general purpose languages. This takes a lot of effort, and it is easier to make mistakes. Instead of writing the type checker in a general purpose language, in this thesis we will systematically derive a type checker from a high-level, declarative specification. This allows for a cleaner implementation that is easy to extend and maintain.

@@

However, this language is not very practical to write actual complex programs in, so to improve this situation we will extend the language with typical features of dependently typed languages. We will provide the following additional contributions:
\begin{itemize}
	\item We show the language is easily extendable, by extending it with booleans among other features (Chapter~\ref{chap:bools})
	\item We show how to add inference of implicit arguments to the implementation (Chapter~\ref{ch:inference})
	\item We show how to add support for inductive datatypes to the implementation (Chapter~\ref{ch:datatypes})
	\item We show how to add support for universes to the implementation (Chapter~\ref{ch:universes})
\end{itemize}

@@

Furthermore, the thesis contains some discussions about the contributions:
\begin{itemize}
	\item We discuss how well semantic code completion works for our language (chapter~\ref{chap:editor-services})
	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi})
	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics})
	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion})
\end{itemize}

@@

The source code for this thesis can be found online \\ \url{https://github.com/JonathanBrouwer/master-thesis}

>>>>

While we keep building more and more complex programming languages, their type checkers are still often written in general purpose languages. This takes a lot of effort, and it is easy to make mistakes. Instead of writing the type checker in a general purpose language, in this thesis we will systematically derive a type checker from a high-level, declarative specification. A declarative specification defines what the semantics of the language are, rather than specifying the imperative steps to type check the language. This allows for a cleaner implementation that is easy to extend and maintain.

@@

However, this language is not very practical to write actual complex programs in, so to improve this situation we will extend the language with typical features of dependently typed languages. We will provide the following additional contributions:
\begin{itemize}
	\item We show the language is easily extendable, by extending it with booleans among other features (chapter~\ref{chap:bools}).
	\item We show how to add inference of implicit arguments to the implementation (chapter~\ref{ch:inference}).
	\item We show how to add support for inductive datatypes to the implementation (chapter~\ref{ch:datatypes}).
	\item We show how to add support for universes to the implementation (chapter~\ref{ch:universes}).
\end{itemize}

@@

Furthermore, the thesis contains some discussions about the contributions:
\begin{itemize}
	\item We discuss how well semantic code completion, which is an editor service provided by Spoofax that completes holes in expressions, works for our language (chapter~\ref{chap:editor-services}).
	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi}).
	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics}).
	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion}).
\end{itemize}

@@

The source code for this thesis can be found online \\ \url{https://github.com/JonathanBrouwer/master-thesis}.

### 2023-04-03 paper-thesis/chapters/03-background-dependent-types.tex
First part of feedback from Jesper

This chapter will explain the concepts behind dependent types. Readers already familiar with Dependent Types are recommended to only skim through this chapter.

@@

A dependent type system is a type system in which types may depend on values. This means it is possible to have a type such as \verb|Vec n|, which is a vector of exactly length \verb|n|. Note that the value of \verb|n| does not have to be known at compile time, it is more powerful than that.

The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example:
\begin{lstlisting}
append : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n -> Vec A (n + n)
\end{lstlisting}
A \verb|Vec A n| is a list with elements of type \verb|A| that is exactly length \verb|n| (an integer, which is a value!). This append function appends two \verb|Vecs| of equal length, returning a \verb|Vec| of length \verb|n + n|. This is an example of a type-level computation.

Dependent types are useful because dependent type systems allow for the Curry-Howard Correspondence~\cite{chc}. Using this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition. Thus if our type system is sound\footnote{The type system we will define in chapter~\ref{chap:baselang} is not sound, we will fix this in chapter~\ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things.

@@

The \textit{Calculus of Constructions} is the language that has all three of these features. Now that we have explained the concepts behind Spoofax and Dependent Types, we will further explain the Calculus of Constructions and implement it in Spoofax in chapter~\ref{chap:baselang}. Then in the chapters after that, we will extend the language with more features.

>>>>

This chapter will explain the concepts behind dependent types. Readers already familiar with dependent types are recommended to only skim through this chapter.

@@

A dependent type system is a type system in which types may depend on values. This means it is possible to have a type such as \verb|Vec n|, which is a vector of exactly length \verb|n|. Note that the value of \verb|n| does not have to be known at compile time, it is therefore more powerful than functions that can be polymorphic over a value.

The biggest advantage of dependent types is that they increase the expressiveness of a type system a lot, for example:
\begin{lstlisting}
append : (A: Set) -> (n : Nat) -> Vec A n -> Vec A n -> Vec A (n + n)
\end{lstlisting}
A \verb|Vec A n| is a list with elements of type \verb|A| that is exactly length \verb|n| (an integer, which is a value!). This append function appends two \verb|Vecs| of equal length, returning a \verb|Vec| of length \verb|n + n|. This is an example of a type-level expression.

In dependent type systems we can use the Curry-Howard Correspondence to prove properties of our code~\cite{chc}. Under this correspondence, types correspond to propositions and a term of a type correspond to a proof of a proposition. Thus if our type system is type-sound\footnote{The type system we will define in chapter~\ref{chap:baselang} is not type-sound, we will fix this by introducing universes in chapter~\ref{ch:universes}.}, that is, we cannot create a term of an empty type, then we can use our type systems to prove things. The type checker will then verify that our proofs are correct.

@@

The \textit{Calculus of Constructions} is the language that has all three of these features. Now that we have explained the concepts behind Spoofax and dependent types, we will further explain the Calculus of Constructions and implement it in Spoofax in chapter~\ref{chap:baselang}. Then in the chapters after that, we will extend the language with more features.

### 2023-04-03 paper-thesis/chapters/10-coc.tex
First part of feedback from Jesper

A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare. Beta reduction is the process of reduction a term to its beta normal form, which is the state where no further beta reductions are possible~\cite{tapl}. It works by matching a term of the form \verb|(\x. b) e|. and substituting \verb|x| in \verb|b| with \verb|e|. Beta reduction applies this rule anywhere in the term, whereas beta head-reduction only applies this rule at the head (outermost expression) of the term, and produces a term in beta-head normal form.

We implemented beta-head reduction using a Krivine abstract machine~\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. This is a strategy under which the leftmost, outermost term is always reduced first~\cite{tapl}. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-subst}.

In conventional dependently typed languages, evaluation is often done using De Bruijn indices.  De Bruijn representation~\cite[Section 6.1]{tapl} uses the distance from a binder to identify a variable. In this representation, alpha equivalence is the same as syntactic equivalence, which can simplify the manipulation of terms. However, we chose to use names rather than De Bruijn indices, because scope graphs work based on names. Using De Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition).

@@

Additionally, we define \verb|betaReduce| which fully beta reduces a term. It works by first calling \verb|betaReduceHead| and then matching on the head, calling \verb|betaReduce| on the sub-expressions of the head recursively.

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $\beq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
	\item $\bred{s_1}{e_1}{e_2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $\scope{\co{sEmpty}}{e}$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

>>>>

A unique requirement for dependently typed languages is beta reduction during type checking, since types may require evaluation to compare. Beta reduction is the process of reduction a term to its beta normal form, which is the state where no further beta reductions are possible~\cite{tapl}. It works by matching a term of the form \verb|(\x. b) e|. and substituting \verb|x| in \verb|b| with \verb|e|. Beta reduction applies this rule everywhere in the term, whereas beta head-reduction only applies this rule at the outermost expression of the term, and produces a term in beta-head normal form, which means the outermost constructor is the same as the outermost constructor of the normal form.

We implemented beta-head reduction using a Krivine abstract machine~\cite{krivine}. The machine can head evaluate lambda expressions with a call-by-name semantics. This is a strategy under which the leftmost, outermost beta-redex is always reduced first~\cite{tapl}. It works by keeping a stack of all arguments that have not been applied yet. This turned out to be the more natural way of expressing this compared to substitution-based evaluation relation, which is an alternative we will discuss in section \ref{sec:coc-subst}.

In conventional dependently typed languages, variables are often represented using de Bruijn indices. De Bruijn representation~\cite[Section 6.1]{tapl} uses the distance from a binder to identify a variable. In this representation, alpha equivalence is the same as syntactic equivalence, which can simplify the manipulation of terms. However, we chose to use names rather than de Bruijn indices, because scope graphs work based on names. Using de Bruijn indices would also prevent us from using editor services that rely on \verb|.ref| annotations (which are Spoofax annotations that declare one name as being a use of another name that is a definition).

@@

Additionally, we define \verb|betaReduce| which fully beta reduces a term. It works by first calling \verb|betaReduceHead| and then matching on the outermost constructor, calling \verb|betaReduce| on the sub-expressions of the outermost constructor recursively.

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|.
	\item $\beq{\scope{s_1}{e_1}}{\scope{s_2}{e_2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|.
	\item $\bhr{\scope{s_1}{e_1}}{\overline{p}}{\scope{s_2}{e_2}}$ is the same as \verb|betaReduceHead((s1, e1), ps) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms}).
	\item $\bred{s_1}{e_1}{e_2}$ is the same as \verb|betaReduce((s1, e1)) == e2|.
	\item $\scope{\co{sEmpty}}{e}$ is the same as $e$ (empty scopes can be left out).
\end{enumerate}

### 2023-04-03 paper-thesis/chapters/30-comparing-with-haskell.tex
First part of feedback from Jesper

\section{Defining the AST}
To implement the calculus of constructions in Haskell, we first define the inductive datatype \verb|Expr|. We chose to define the language using De Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones. 
\begin{lstlisting}
data Expr =
	Type
	| Let Expr Expr
	| Var Int
	| FnType Expr Expr
	| FnConstruct Expr Expr
	| FnDestruct Expr Expr
\end{lstlisting}
There is a constructor for each constructor in our language. The difference with the implementation in Spoofax is that the \verb|Let|, \verb|FnType|, and \verb|FnConstruct| no longer store an identifier, instead they introduce a new binding into their body, which are accessible by their De Bruijn indices.

@@

Next, we need to decide how we want to store the environment. We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter~\ref{chap:baselang}. We will use the same names, and store the environment as a list, with the head of the list representing De Bruijn index 0. Finally, we define the type of a scoped expression \verb|SExpr|, as a tuple of an environment and an expression.
\begin{lstlisting}
data EnvEntry = NType Expr | NSubst SExpr
type Env = [EnvEntry]
type SExpr = (Env, Expr)
\end{lstlisting}

>>>>

\section{Defining the AST}
To implement the calculus of constructions in Haskell, we first define the inductive datatype \verb|Expr|. We chose to define the language using de Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones. 
\begin{lstlisting}
data Expr =
	Type
	| Let Expr Expr
	| Var Int
	| FnType Expr Expr
	| FnConstruct Expr Expr
	| FnDestruct Expr Expr
\end{lstlisting}
There is a constructor for each constructor in our language. The difference with the implementation in Spoofax is that the \verb|Let|, \verb|FnType|, and \verb|FnConstruct| no longer store an identifier, instead they introduce a new binding into their body, which are accessible by their de Bruijn indices.

@@

Next, we need to decide how we want to store the environment. We still need to store both function arguments and substitutions, which we called \verb|NType| and \verb|NSubst| in chapter~\ref{chap:baselang}. We will use the same names, and store the environment as a list, with the head of the list representing de Bruijn index 0. Finally, we define the type of a scoped expression \verb|SExpr|, as a tuple of an environment and an expression.
\begin{lstlisting}
data EnvEntry = NType Expr | NSubst SExpr
type Env = [EnvEntry]
type SExpr = (Env, Expr)
\end{lstlisting}
